A Global Picture of Food Insecurity and Hunger

### 2020-10-30 docs/modeling_fies.tex
almost done with paper; improve visualizations

@@

Examining the rate of food insecurity in a population, rather than the total number of food insecure people, our model finds the most progress has been made in East \& Southeast Asia, while Central \& Southern Asia is expected to see improvements in food insecurity throughout the 2020s.  Other poor and middle income areas, including much of Africa, the Middle East and Latin America, have made only slight progress and will continue to make slight progress between 2010 and 2030.  Finally, in high-income countries in Europe and North America, rates of people in at least moderate food insecurity move little, and rates of severe food insecurity are expected to increase slightly.

@@

\section{Discussion}
This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts out to the year 2030.  We found that, while food insecurity has been increasing and has likely worsened in 2020 with economic downturn associated with the novel coronavirus, our models predict that number of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in south and east Asia.  Nevertheless, many regions, most notably sub-Saharan Africa, are expected to see overall increases in the number of food insecure people. %Add here comparison with SOFI, but emphasize that discrepancies may also come from measuring different things

%Homi says delete this para
These results contrast with other projections that have been made for While this is the first comprehensive projection of indicator 2.1.2 for SDG 2, projections for other indicators for that goal have been made.  The most recent State of Food Insecurity (SOFI) report by the FAO projected indicator 2.1.1, the prevalence of undernourishment, and found that that, much like food insecurity, undernourishment had been increasing throughout the 2010s \citep{sofi2020}.  The SOFI report also modeled the prevalence of undernourishment in 2030 based on recent trends, giving more weight to recent years, and projected that undernourishment will increase by 2030 \citep[Part 1, Page 11]{sofi2020}.  Using a similar method for modeling the prevalence of stunting (indicator 2.2.1), the SOFI reported that the rate of stunting would decrease by 2030 \citep[Part 1, Page 30]{sofi2020}.

The fact that the SOFI expects increases in undernourishment while we predict decreases in food insecurity is likely due to our differing modeling approaches. While the SOFI report extrapolated recent trends in undernourishment, we modeled food insecurity based on projections of factors like GDP per capita, population, and the prevalence of stunting.  We found that, in a random forest model, these predictors were very successful at predicting recent rates of both moderate-to-severe and severe food insecurity.

%Add section about the drivers of food insecurity, as well as how they largely account for the outcomes observed.

While we found that food insecurity will decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, only expects the number of people experiencing moderate food insecurity to fall by 12\% and the number of people experiencing severe food insecurity to fall by 20\%.  This still leaves billions of people eating less than they should and nearly half a billion people going entire days without eating a decade from now, a number that falls far short of the SDG 2 goal of ``zero hunger."  Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, or obesity.  Thus, while expected trends in correlates of food security give cause for optimism, there is still significant work to be done in pursuit of SDG 2.

@@

\subsection{Modeling}
We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.  We used random forest regressions, as this approach performs well in the presence of non-linearities and important interactions among predictor variables \citep{hastie2009elements}, while providing more model interpretability than other machine learning approaches \citep{ishwaran2007variable}.  A random forest regression involves generating a large number of decision trees, each under slightly different conditions and using different sub-samples of the data, and then aggregating the predictions of the decision trees.

Because the outcome variables in our models, the rate of moderate or severe food insecurity, is a bounded outcome, we first used a logistic transformation to extend the range from $[0,1]$ to $(-\infty, \infty)$.  Then, after estimating the model, we used the inverse logit to convert the unbounded model predictions to a rate between 0 and 1.

To get the best possible model performance, it is necessary to tune the hyperparamters, which control the process in which the random forest algorithm is run.  Key hyperparameters include the number of decision trees to generate, the number of variables randomly selected as candidates for splitting a node, and the average number of observations in a leaf node.  We calculated the Out-Of-Bag (OOB) error for different combinations of hyperparameters for the moderate and severe models. By choosing the combinations of hyperparameters with the smallest OOB error we found the optimal combination for both models. We found that 5000 trees are more than sufficient for the error rates to converge (See Fig. \ref{fig:rf_error}).

%Talk more about validation: excluding HICs, how many runs, etc
We observed for the moderate as well as the severe model an in-sample mean squared error (MSE) between 0.002-0.005, with a R-squared (${R}^2$) value of 0.999 (See Fig. \ref{fig:rf_in-sample}). To test the out-of-sample prediction performance we performed ten-fold cross-validation. First, we partitioned the data into ten equal sized subsamples. Once partitioned, we trained the RF regression models on nine subsamples and tested the models on the remaining one. This process gets repeated ten times, with each subsample being used once for testing. We then calculated the average MSE and ${R}^2$ over all ten out-of-sample predictions. Similarly to the in-sample model validation metrics, the average out-of-sample MSE is between 0.008-0.016, with an average ${R}^2$ of around 0.99 for both models (See Fig. \ref{fig:rf_out-sample}). Besides using the full set of countries to validate the out-of-sample performance, we halved the sample size by excluding high-income countries (HIC) (according to the World Bank definition, REFERENCE NEEDED) and performed ten-fold cross-validation for this limited set of countries. We found that even when excluding HIC the average out-of-sample MSE is between 0.013-0.023, with an average ${R}^2$ of around 0.98 for both models.

After fitting the models, we examined the importance of the individual covariates in explaining rates of moderate and severe food insecurity. By training the models with all covariates and predicting rates with and without a individual variable we calculated the average prediction error of omitting that variable. Using the highest average prediction error we ranked the importance of the variables and found that for both, the moderate-to-severe and severe model, the Poverty Headcount Index, Stunting, GDP Per Capita and the Gini Coefficient are the most important variables for explaining differences in rates of moderate-to-severe and severe food insecurity (See Fig. \ref{fig:rf_vimp}).

@@

>>>>

% Ask Carlo - has the FAO made predictions of the FIES yet?

@@

Examining the rate of food insecurity in a population, rather than the total number of food insecure people, our model finds the most progress has been made in East \& Southeast Asia, while Central \& Southern Asia is expected to see improvements in food insecurity throughout the 2020s.  Other poor and middle income areas, including much of Africa, the Middle East and Latin America, have made only slight progress and will continue to make slight progress between 2010 and 2030.  In high-income countries in Europe and North America, rates of people in at least moderate food insecurity move little, and rates of severe food insecurity are expected to increase slightly.  Overall, the world has made has made steady but slow progress on reducing the rate of food insecurity.

@@

\section{Discussion}
This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts out to the year 2030.  We found that, while food insecurity has been increasing and has likely worsened in 2020 with economic downturn associated with the novel coronavirus, our models predict that number of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in south and east Asia.  Nevertheless, many regions, most notably sub-Saharan Africa, are expected to see overall increases in the number of food insecure people.

Our variable importance metrics show that economic conditions are the largest driver of global variation in rates of food insecurity.  For both moderate and severe food insecurity, the poverty headcount index, the Gini coefficient, and GDP per capita were major predictors of the rate of food insecurity.  The rate of stunting, an indicator of both sanitary conditions and chronic, long-term hunger was also an important predictor of food insecurity.  However, other indicators of sanitary conditions, such as the rates of wasting and malaria mortality were less important predictors of food insecurity.  Geographic and climate factors, such as urban percentage, topographic ruggedness and mean annual temperature and precipitation were not important predictors of food insecurity at a global scale.  The fact that our models had such high accuracy indicates that these variables alone are sufficient to predict food insecurity globally across countries, although other variables may be more important for explaining variation at more local scales.

These results contrast with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.  This is likely due in part to our differing modeling approaches. While the SOFI report extrapolated recent trends in undernourishment and stunting, we modeled food insecurity based on interactions between projections in factors like GDP per capita, population, and the prevalence of stunting. Additionally, our diverging predictions for different SDG 2 indicators are not necessarily mutually exclusive: it is theoretically possible for undernourishment to increase while food insecurity, as measured by the FIES, decreases.  The discrepancies between our projections illustrate both how hunger is a complex phenomenon that is resistant to simple quantification, as well as the challenges in forecasting the future, where different modeling approaches will yield different predictions.

When making predictions in countries where we did not have primary data, we are assuming that the relationship between variables like poverty, stunting, inequality, and hunger are the same in all locations.  We were able to test whether this was the case using data from countries that reported their aggregate rates of food insecurity but did not release microdata, and we found that our model largely proved well in those cases, with an accuracy of (ACCURACY HERE).  Additionally, when making predictions into the future, we are assuming that long-term patterns of demographic change, urbanization, and development will maintain their long-held trajectories for the next decade.  At a global scale, this is almost certainly the case: it is highly unlikely that rates of fertility or economic growth with shift suddenly and globally.  Nevertheless, at more local scales, sudden crises can lead to severe and unforeseeable increases in food insecurity, as crises in the previous decade in Yemen, Syria and Venezuela have shown.  Thus, while our model's predictions at a global scale are very likely to bear out, there will probably be individual countries and areas where our forecasts end up being off.

While we found that food insecurity will decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, only expects the number of people experiencing moderate food insecurity to fall by 12\% and the number of people experiencing severe food insecurity to fall by 20\%.  This still leaves billions of people eating less than they should and nearly half a billion people going entire days without eating a decade from now, a number that falls far short of the SDG 2 goal of ending hunger.  Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, or obesity.  Thus, while expected trends in correlates of food security give cause for optimism, there is still significant work to be done in pursuit of SDG 2.

@@

\subsection{Modeling}
We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.  We used random forest regressions, as this approach performs well in the presence of non-linearities and important interactions among predictor variables \citep{hastie2009elements}.  A random forest regression involves generating a large number of decision trees, each under slightly different conditions and using different sub-samples of the data, and then aggregating the predictions of the decision trees.

Because the outcome variables in our models, the rate of moderate or severe food insecurity, are a bounded outcome, we first used a logistic transformation to extend the range from $[0,1]$ to $(-\infty, \infty)$.  Then, after estimating the model, we used the inverse logit to convert the unbounded model predictions to a rate between 0 and 1.

To get the best possible model performance, it is necessary to tune the hyperparamters, which control the process in which the random forest algorithm is run.  Key hyperparameters include the number of decision trees to generate, the number of variables randomly selected as candidates for splitting a node, and the average number of observations in a leaf node.  We calculated the Out-Of-Bag (OOB) error for different combinations of hyperparameters for the moderate and severe models. By choosing the combinations of hyperparameters with the smallest OOB error we found the optimal combination for both models. We found that 5000 trees are more than sufficient for the error rates to converge.

After fitting the models, we examined the importance of the individual covariates in explaining rates of moderate and severe food insecurity. We used a common method of permuting each individual variable and re-running the regressions, and then determining the difference in Mean Average Error (MAE) between the regression \citep{ishwaran2007variable}.  Similarly to model predictions, we transformed the error term with the inverse logit so that the errors would be on the scale of the outcome variable and therefore more interpretable.

\subsection{Model Validation}
We observed a MAE of 0.0029 for our model of severe food insecurity ($R^2$ of 0.9989) and an MAE 0.0054 for our model of moderate food insecurity ($R^2$ of 0.9990).  This means that, on average, our predictions were only of by 0.29\% and 0.54\%, respectively, for the subnational areas in countries where we had microdata.

To further validate the models, we aggregated our predictions to the country level to compare with national-level reports by the FAO.  These included countries which reported the rate of people over the thresholds for moderate and severe food insecurity, but did not release raw microdata, thus providing an truly out-of-sample means of testing our model.  Among these 21 countries, our model predicted the rate of at least moderate food insecurity with a MAE of 0.0961 ($R^2$ = 0.8094), and severe food insecurity with a MAE of 0.0718 ($R^2$ = 0.7782).

@@

To test the out-of-sample prediction performance we performed ten-fold cross-validation. First, we partitioned the data into ten equal sized subsamples. Once partitioned, we trained the RF regression models on nine subsamples and tested the models on the remaining one. This process gets repeated ten times, with each subsample being used once for testing. We then calculated the average MSE and ${R}^2$ over all ten out-of-sample predictions. Similarly to the in-sample model validation metrics, the average out-of-sample MSE is between 0.008-0.016, with an average ${R}^2$ of around 0.99 for both models (See Fig. \ref{fig:rf_out-sample}). Besides using the full set of countries to validate the out-of-sample performance, we halved the sample size by excluding high-income countries (HIC) (according to the World Bank definition, REFERENCE NEEDED) and performed ten-fold cross-validation for this limited set of countries. We found that even when excluding HIC the average out-of-sample MSE is between 0.013-0.023, with an average ${R}^2$ of around 0.98 for both models.
